Summary Statistics

### 2025-03-04 extensive_tests/draft-sumstat-vignette.Rmd
suggest a few changes in vignettes on summary stats

---
title: "Summary Statistics"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Summary Statistics with MethEvolSIM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

This document presents summary statistics related to the evolution of methylation 
states across different genomic structures. The statistics include mean 
frequencies, standard deviations, neighbor correlations of methylation states in 
island and non-island regions. Additionally, for analysis comparing different 
samples (comparison of tree tips), the included summary statistics compute
the mean per-site frequency of methylation changes and the Fitch estimate for
the minimum number of changes per islands leading to different global methylation
states.

The provided set of functions allows the computation of summary statistics for 
methylation data in a genomic region with structural categorization. In this 
vignette, we will refer to these structures as **islands** and 
**non-island structures**.

@@

```{r}
# 2 tip / sample / replicate
sample_n <- 2
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(
  list(c(.5, .5, 0, 0, 0, .5), c(.5, 0, 0, .5), c(.5, .5, 0), c(0, 0, .5)), # tip 1
  list(c(0, .5, .5, 1, 1, .5), c(1, .5, 1, .5), c(0, .5, .5), c(1, .5, 1)) # tip 2
  )
get_islandMeanFreqP(index_islands, data, sample_n)
get_nonislandMeanFreqP(index_nonislands, data, sample_n)
get_islandMeanFreqM(index_islands, data, sample_n)
get_nonislandMeanFreqM(index_nonislands, data, sample_n)
```

@@

For a given cherry $(t_1, t_2)$, we compute the proportion of sites with 
differing methylation states between the two tips.

@@

- `freqSites_cherryMethDiff(tree,data)`. The function first validates the tree 
structure and extracts pairwise distances between cherry tips 
with `get_cherryDist`. It then validates the input data and counts half and full 
methylation differences for each cherry at each structure using 
`countSites_cherryMethDiff`. Finally, it normalizes these (half and full) counts 
by the number of sites per structure to compute frequencies. .

- `get_siteFChange_cherry(tree,data)`. Uses `freqSites_cherryMethDiff(tree,data)` 
and then computes the frequency of sites with any change of methylation state 
(full or half) for each cherry at each genomic structure.

@@

Given a phylogenetic tree $T$ with tips corresponding to sampled species or 
individuals, we aim to estimate the minimum number of state changes required to 
explain the observed distribution of island methylation states across the tips. 
This is achieved using Fitch parsimony, a dynamic programming approach 
implemented in `compute_fitch(meth, tree)` with the following steps:

- Each tip $t$ is assigned the observed methylation state for each island $i$.
- Internal nodes recursively inherit possible state sets by computing the 
intersection of the state sets from their child nodes.
- If the intersection is empty (i.e., no common state between child nodes), the 
union of both state sets is assigned, and the number of changes is incremented.

>>>>

---
title: "Summary Statistics"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: ../vignettes/references.bib
vignette: >
  %\VignetteIndexEntry{Summary Statistics with MethEvolSIM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

Here we present summary statistics that can be used,
e.g. in an Approximate Bayesian Computation method [@csillery2012abc],
to estimate parameters of the methylation dynamics model of MethEvolSIM.
The statistics include mean 
frequencies, standard deviations, neighbor correlations of methylation states in 
island and non-island regions.
Additionally, for analyses comparing different 
samples (comparison of tree tips), some of the summary statistics are based
on the mean per-site frequency of methylation changes and
the minimum number of changes per islands leading to different global methylation
states, calculated with the Fitch algorithm [@fitch1971toward].

The provided set of functions allows the computation of summary statistics for 
methylation data in a genomic region with structural categorization
into two types of structures.
In this  vignette, we will refer to these structures as **islands** and 
**non-island structures**.

@@

```{r}
# 2 tip / sample / replicate
sample_n <- 2
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(
  list(c(.5, .5, 0, 0, 0, .5), c(.5, 0, 0, .5), c(.5, .5, 0), c(0, 0, .5)), # tip 1
  list(c(0, .5, .5, 1, 1, .5), c(1, .5, 1, .5), c(0, .5, .5), c(1, .5, 1)) # tip 2
  )
get_islandMeanFreqP(index_islands, data, sample_n)
get_nonislandMeanFreqP(index_nonislands, data, sample_n)
get_islandMeanFreqM(index_islands, data, sample_n)
get_nonislandMeanFreqM(index_nonislands, data, sample_n)
```
Note that in the calculation of the means the relative frequencies $x_{i, t}$, 
$y_{j, t}$, $w_{i, t}$ and $z_{j, t}$ are **not** weighted by the lengths of 
island $i$ or non-island $j$.

@@

For a given cherry $(t_1, t_2)$, that is two tips $t_1$ and $t_2$ that are direct
offspring of the same internal node, we compute the proportion of sites with 
differing methylation states between the two tips.

@@

- `freqSites_cherryMethDiff(tree,data)`. The function first validates the tree 
structure and extracts pairwise distances between cherry tips 
with `get_cherryDist`. It then validates the input data and counts half and full 
methylation differences for each cherry at each structure using 
`countSites_cherryMethDiff`. Finally, it normalizes these (half and full) counts 
by the number of sites per structure to compute frequencies.

- `get_siteFChange_cherry(tree,data)` uses `freqSites_cherryMethDiff(tree,data)` 
and then computes the frequency of sites with any change of methylation state 
(full or half) for each cherry at each genomic structure.

@@

Given a phylogenetic tree $T$ with tips corresponding to sampled species or 
individuals, we aim to estimate the minimum number of state changes required to 
explain the observed distribution of island methylation states across the tips. 
This is achieved using the Fitch algorithm [@fitch1971toward] as
implemented in `compute_fitch(meth, tree)`.
